Unsupervised Subclassification of Spectroscopic Surveys by Dimensionality Reduction: DESI White Dwarfs

### 2024-08-16 src/tex/manuscript.tex
Third draft

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[
    Unsupervised WD classification
]{
    Unsupervised White Dwarf Spectral Classification by Dimensionality Reduction
}

@@

% Abstract of the paper
\begin{abstract}
As a new generation of large-sky spectroscopic surveys comes online, the enormous data volume poses challenges in classifying spectra.
Although machine-learning-based spectral classifiers exist, they are usually trained in a supervised way, and their performance thus relies heavily on training data.
We explore the use of dimensionality reduction to transform the high-dimensional space of white dwarf spectra, from the Dark Energy Spectroscopic Instrument's Early Data Release, into a two-dimensional map.
This provides a new view of the dataset which groups together white dwarfs of different spectral classes.
By cropping the spectra to windows around particular spectral lines, prior knowledge about the wavelengths of important spectral features can be exploited to isolate particular spectral classes.
Although dimensionality reduction is a fully unsupervised technique, it can also be used in a supervised way, allowing spectra from other sources to be classified by visualising their similarity to the spectra in the dataset.
With upcoming surveys promising tens of millions of spectra, this work highlights the potential for unsupervised techniques as efficient means of classification and dataset visualisation.
\end{abstract}

@@

White dwarfs (WDs) are the final state of the $\gtrsim97$ per cent of stars with a zero-age mass of less than $9$--$12\,\mathrm{M}_\odot$ \citep{lauffer18, althaus10, althaus21}.
Without internal nuclear fusion, these stellar cinders gradually cool on well-characterised timescales, making them important tracers of the evolution and assembly of the Galaxy (e.g., \citealt{winget87, tremblay14}).
The spectrum of a white dwarf usually deviates from that of a black body due to absorbing species in its atmosphere.
With a typical mass of $\sim0.6\,\mathrm{M}_\odot$ and radius of $\sim \mathrm{R}_E$, the extreme gravitational field gives WDs a strongly stratified structure.
Metals diffuse down through WD atmospheres on timescales much shorter than their cooling age \citep{schatzman45, paquette86, koester09, wyatt14}, leaving thin atmospheres dominated by H and/or He.
Most WD spectra show absorption features due to these light elements, and a WD is classified respectively as DA or DB if spectral features of H or He are present.
As these WDs cool below particular temperatures, atomic transitions are no longer excited ($\lesssim5000\,\text{K}$ for H; $\lesssim 11000\,\text{K}$ for He) and the absorption lines gradually fade away, giving a featureless spectrum (DC).
In certain WDs, the convection zone reaches deep into the interior, dredging up carbon from the inert core and imprinting molecular \ce{C2} features in the spectrum (DQ; \citealt{fontaine84, koester82, koester20, blouin23}).
In other cases, instabilities in remnant planetary systems can cause planetesimals to be tidally disintegrated and accreted by WDs, lending metal absorption features to the spectrum \citep{bonsor11, frewen14, mustill18, maldonado20}.
These so-called \textit{polluted} WDs (DZ) provide rare insights into the bulk composition and geology of rocky exoplanets.

Following the release of the DESI EDR \citep{desiedr}, the spectra of 3673 WD candidates targeted in this data release were visually inspected and classified by \citet{manser24}, spectroscopically confirming 2706 WDs, of which 1400 had not been identified by previous spectroscopic surveys.
In particular, 152 polluted WDs were found, of which 121 were newly discovered or previously not classified as polluted: the higher resolution of DESI spectra enables the identification of metal lines not visible in lower-resolution spectra.
However, visual classification requires significant amounts of expert time, with substantial increases in target numbers in the future ($\sim70\,000$ WD candidates from the DESI first data release; \citealt{cooper23}), increasing the benefits of automation.

One class of methods for automatically classifying and analysing WD spectra that has achieved some success is supervised machine learning (for example, \citealt{yang20, tan23, garciazamora23, vincent23, vincent24}).
However, a drawback of supervised machine learning techniques is the reliance on the training set used to train models.
The `imbalanced learning problem' describes the difficulties of training supervised classifiers on datasets with large class imbalances \citep{he09, johnson19}.
Such class imbalances are inevitable in WD datasets as some spectral types are much rarer than others.
Additionally, incorrect labels in the training set limit the accuracy of the resulting model (e.g., \citealt{frenay14}).
Unsupervised procedures such as dimensionality reduction do not rely on an external training set; they merely reveal structures present within the dataset itself.

In this work, we apply dimensionality reduction -- specifically, $t$SNE -- to white dwarf spectra from the DESI EDR, exploring the method's success in classifying and characterising WD spectra in an unsupervised way.
We emphasise that these methods could equally be applied to sets of main-sequence stars, quasars, or galaxies, all of which are targeted by the aforementioned spectroscopic surveys.
The paper proceeds as follows.
In Section~\ref{sec:methods}, we describe the dataset and $t$SNE method in more detail.
In Section~\ref{sec:results}, we demonstrate this method's ability to identify clusters and sequences in the DESI EDR WD catalogue, a way to incorporate prior knowledge about the location of distinctive spectral lines, and a method of estimating the classification of spectra external to the catalogue.
Section~\ref{sec:discussion} discusses the application of this method to the bulk classification of large spectroscopic WD datasets.
Section~\ref{sec:conclusions} summarises our work.

@@

The DESI EDR contains $N=3673$ candidate WDs from the catalogue compiled by \citet{gentilefusillo19} whose exposures include a median signal-to-noise ratio $>0.5$ in at least one of DESI's three spectral arms \citep{manser24}.
These publicly available\footnote{
    \url{https://data.desi.lbl.gov/public/index.html}
} exposures were obtained and restacked.
The spectra span $3600$--$9824\,\text{\AA}$, with a wavelength spacing of $\Delta \lambda = 0.8\,\text{\AA}$, giving $D=7781$ flux bins for each spectrum\footnote{
    Note that the full-width-half-maximum line width of the DESI spectrographs is $\approx1.8\,\text{\AA}$ \citep{abareshi22}; these data are therefore slightly oversampled.
}.

@@

A spectrum can be represented by a high-dimensional vector, where each component corresponds to the flux in a given wavelength bin.
Spectra produced by the DESI pipeline have 7781 flux bins, so each spectrum can be mapped to a 7781-dimensional vector.
The vectors are not uniformly-distributed in $\mathbb{R}^{7781}$;
spectra with shared spectral features are clustered together, as their vector components corresponding to wavelength bins near shared spectral features will take similar values.
For example, component 3703 corresponds to a wavelength of $6562.4\,\text{\AA}$, very close to the H$\,\alpha$ line.
The normalised spectra of DA WDs will have lower values in their 3703rd component than spectra \textit{without} any H absorption features
DAs will hence -- at least in this component, and by extension other components -- be close together.
This inherent structure in the data means that each point in the dataset lies on or close to a submanifold of the data space with much lower dimensionality. DR techniques seek to find a low- (usually two-) dimensional representation of the structure of the high-dimensional dataset.

@@

Before applying DR, it is of paramount importance to preprocess the spectra, as shown in Fig.~\ref{fig:preprocessing}.
Some spectra show large spikes, due to cosmic rays or detector artefacts.
Interpreted as vectors, this corresponds to some components of the vector being very large, and as a result the vector would be very far away in data space from where it would be without the artefacts.
These spikes usually correspond to pixels with very low signal-to-noise: pixels with $\mathrm{S}/\mathrm{N}<0.2$ are linearly interpolated.
Fig.~\ref{fig:preprocessing} shows that major, though not all, artefacts are removed by this step.
Overzealous removal of artefacts could remove genuine spectral features, such as the sharp emission lines of cataclysmic variables (see Section~\ref{sec:CVs}): we found that the interpolation described improved the effectiveness of the subsequent dimensionality reduction as compared to, say, median boxcar smoothing.

@@

Additionally, the spectra show a range of absolute scales: the median fluxes span three orders of magnitude.
As we are interested in intrinsic properties of the WDs in this sample, rather than their distance from Earth, the spectra are normalised to zero mean and unit variance (see Fig.~\ref{fig:preprocessing}).

In its simplest form, dimensionality reduction treats all components of the vectors equally.
However, some wavelengths are naturally of more significance than others, namely those of atomic absorption and emission features.
To `focus' the technique on a particular spectral line, we crop the spectrum to a window containing that line (as well as some continuum) before applying DR.
This additional preprocessing step is explored in Section~\ref{sec:zoom}.

@@

Dimensionality reduction was applied to the DESI EDR WD catalogue, using $t$SNE.
The dimensionality reduction takes an average of $5.5\,\text{s}$ on an Intel $3.60\,\text{GHz}$ CPU.
The resulting embedding is shown in Fig.~\ref{fig:full_spectra}(a).
Each point in this embedding corresponds to an individual spectrum, and
several clusters and sequences appear.
There is a long V-shaped sequence stretching from the left, to the bottom, to the upper right of the embedding; it turns off into a cluster towards the centre right.
A shorter second sequence extends from the top of the embedding towards the centre.
Several strings and clumps are found in between and around these two sequences.

The nature of these structures is elucidated in Fig.~\ref{fig:full_spectra}(b) and (c), showing respectively the primary spectral class (acc. \citealt{manser24}) and effective temperature (acc. H-atmosphere model fitting of \citealt{gentilefusillo19}).
Several interesting features include:
\begin{itemize}
\item
The V-shaped sequence consists of primarily DA white dwarfs;
\item
The second sequence in the upper part of the frame contains DBs, DZs, DQs, and DCs;
\item
DZs are found in various places\footnote{
    \citet{kao24} also find that DZs with different characteristics (primarily $\Teff$) are scattered by dimensionality reduction into multiple regions.
}: (i) an island near to the second sequence, (ii) towards the cool end of the DA sequence, (iii) amongst the second sequence;
\item
Extragalactic sources (crosses) are mostly located in strings and small clumps around the second sequence;
\item
The majority of main-sequence stars (black star icons) are found towards the centre right, though their cluster merges somewhat with the cool end of the DA sequence.
A few are also scattered elsewhere;
\item
The WDs in the DA sequence have been sorted approximately according to effective temperature, with hotter WDs towards the left of the sequence\footnote{
    This trend is also apparent in the second sequence, but because many of these are DBs and DQs, and the temperatures shown here are fitted assuming H-atmosphere models, the trend is less meaningful.
    However, the trend is borne out in the fitted temperatures of He-atmosphere models.
};
\end{itemize}

@@

We emphasise that although the data are in this case labelled by class identified by visual inspection, the method would nonetheless provide a valuable analysis if the classes were \textit{a priori} unknown, as will be the case with upcoming spectroscopic surveys.
For example, it is well-known that approximately 80 per cent of WDs are DAs.
As such, when applying DR to a WD dataset one would implicitly identify the largest cluster as mostly DAs.

That DR is capable of separating different classes of WD into groups in this way is ultimately due to similarities between spectra of the same spectral class, and differences between different classes.
Expressed as vectors, DA spectra (say) are much closer to vectors of other DA spectra of similar H abundance and effective temperature, than to DB spectra (say).
Note that this would not be true if the spectra had not been preprocessed (see Section~\ref{sec:preprocessing}): closer WDs would have brighter spectra than more distant WDs of the same spectral type, and hence spectral vectors with larger magnitudes.
The spectral vectors would therefore be very far away from each other, and hence embedded far away from each other, despite being similar astrophysical objects.
The same would be true of spectra with very large artefacts.
The preprocessing procedure is thus crucial in removing any irrelevant aspects of the data.

The H-atmosphere model fitting of \citet{gentilefusillo19} suggests that the WDs in the V-shaped sequence transition smoothly from $\Teff \gtrsim 80000\,\text{K}$ through to $\Teff \lesssim 4000\,\text{K}$.
We see that the well-sampled variation that must exist in the tilt of the spectral black-body continuum is preserved in two dimensions.
The hottest WD in this sequence has a very similar spectrum to the second-hottest, so the two are embedded close together; the second-hottest has a similar spectrum to the third-hottest; etc.
Anomalies of intermediate temperature at opposite ends of the sequence (see Fig.~\ref{fig:full_spectra}(c)) are main-sequence stars or other objects erroneously assigned a WD temperature.
That $\Teff$ ``spans'' the embedding identifies this parameter as that which describes a spectrum to first order: the parameter which primarily determines the spectrum is the temperature.
If $\Teff$ were not of interest, one could include preprocessing steps to erase any temperature information in the spectra (e.g., dividing by the best-fitting black body) before applying DR.
Unable to pick up on this information, DR would then arrange the embedding according to second-order parameters, such as hydrogen abundance.
In such a scenario, the spectrum of a very hot DA would appear similar to that of a very cool DA and the sequence would thus be collapsed in the embedding.

The transitions undergone by WDs as they gradually cool are also borne out in the spectral classifications of the embedding, shown in Fig.~\ref{fig:full_spectra}(b), though some subtleties emerge as a result of the classification system itself.
For example, as the temperature of a DA falls below around $\Teff\approx 5000\,\text{K}$, hydrogen transitions are no longer excited, the Balmer lines gradually fade away, and the DA becomes a DC.
However, the transition is gradual: the Balmer lines do not suddenly disappear below some sharp temperature cutoff.
As such, the distinction between DA and DC is a fuzzy one, and whether a cool hydrogen-atmosphere WD spectrum is visually classified as DA or DC depends whether the classifier can subjectively identify H features against noise.
This task is not trivial, but we argue that it is ultimately arbitrary: there is negligible physical difference between a DA with very weak Balmer lines, and a hydrogen-atmosphere DC.

@@

Similar trends are borne out in the secondary sequence, which shows DBs at the top of the embedding transitioning into a mixture of DCs and DQs.
The transition from DB to DC has a similar physical origin to the DA-to-DC transition discussed in the previous paragraph; as expected this transition occurs at a higher temperature as shown in Fig.~\ref{fig:full_spectra}(c).
The DC transition temperatures appear roughly in line with theoretical predictions of $\approx 5000\,\text{K}$ and $\approx 11000\,\text{K}$ for H- and He-dominated atmospheres respectively.
Aside from forming a DC, a cooling He-atmosphere WD may begin dredging carbon from the core, giving rise to C$_2$ Swan bands in the spectrum, which is then classified DQ \citep{fontaine84}.
As this cooling occurs, He transitions fail to be excited and the He lines of a DB fade away.
As with Balmer lines, He lines and Swan bands can be arbitrarily weak, creating a degree of blurriness between DBs, DQs, and DCs.
That these three spectral types are all projected together in the second sequence at the top of the embedding reflects the smooth transition between them.
Indeed, Fig.~\ref{fig:full_spectra}(c) shows that the DCs in this part of the embedding have temperatures\footnote{
    Although the temperatures shown in Fig.~\ref{fig:full_spectra}(c) are based on H-atmosphere models, the He-atmosphere model temperatures show the same results.
} well in excess of the $\approx 5000\,\text{K}$ DA-to-DC transition.
These DCs must therefore have He-dominated atmospheres; if they were H-dominated then Balmer features would be visible.
These observations therefore suggest that the two main clusters in the embedding correspond to H- and He-dominated atmospheres.

Briefly, we discuss the presence of a small number of DAs in the cool end of the He-atmosphere sequence (see Fig.~\ref{fig:full_spectra}(b), just above centre), of which there are 43.
Of these, 24 are classified by \citet{manser24} as DAH, DAe, DAP, or some combination of these secondary classifications.
We believe that these are found in the He-atmosphere sequence as the Swan bands of the DQs are mimicked by the Zeeman-split Balmer features of a DAH, emission features of a DAe, or other unidentified features in DAPs.
The remaining 19 DAs here are either pure DAs, DABs, DAZs, or some combination of these.
These DAs are invariably warm ($\Teff \gtrsim 7000\,\text{K}$) and have weak Balmer lines, making them appear similar to DCs of a similar temperature.
It is thus understandable that these DAs are projected near to these warm (thus He-atmosphere) DCs.

It is somewhat surprising that there is a gap between the two sequences.
Based on the observations outlined above, this region should be populated with DCs of 
%%% FIXME Do H eq width fitting

The automated nature of DR can protect against human error in the visual classification of WD spectra.
Fig.~\ref{fig:DZisland} shows a zoom-in towards the top of the embedding around an island populated mostly by WDs with primary classification DZ.
According to the extensive visual classifications of \citet{manser24}, this island also contains a DBZ (J133305.34+325400.11) and a DC (J160711.86+532157.65).
These objects appear to have been grouped together owing to the presence of Ca~H and K lines in all the spectra, including the putative DC.
The presence of Ca absorption lines in the DC's spectrum suggests that this WD is in fact a DZ; indeed \citet{kleinman13} classify this source as a DZ based on its SDSS DR7 spectrum.
The bottom panel of Figure~\ref{fig:DZisland} shows the absorption lines to be quite weak, and would probably not have been noticeable without zooming in.
While large visual classification campaigns are reliable, this illustrates that errors are nonetheless possible, and that such errors can be easily identified with automated methods such as DR.

@@

As mentioned in Section~\ref{sec:preprocessing}, DR treats all components of a vector -- i.e., all wavelengths -- equally, but wavelengths including spectral features are of particular interest.
In this subsection we explore the use of cropping the spectrum, to a window around some particular spectral line, to see if DR can better separate individual spectral classes from the rest.

@@

It is clear that dimensionality reduction alone is not capable of classifying the spectra of all white dwarf candidates.
Though the technique is quickly able to classify a large fraction of the dataset, particularly when focused on a wavelength range around some distinctive spectral features, we outline drawbacks to DR in this subsection.

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[
    Unsupervised Subclassification of Spectroscopic Surveys
]{
    Unsupervised Subclassification of Spectroscopic Surveys by Dimensionality Reduction: DESI White Dwarfs
}

@@

% Abstract of the paper
\begin{abstract}
As a new generation of large-sky spectroscopic surveys comes online, the enormous data volume poses challenges in classifying spectra.
Although machine-learning-based spectral classifiers exist, they are usually trained in a supervised way, and their performance thus relies heavily on training data.
We outline the use of dimensionality reduction to transform the high-dimensional space of white dwarf candidate spectra from the Dark Energy Spectroscopic Instrument's Early Data Release into a two-dimensional map.
This provides a new view of the dataset which groups together white dwarfs of different types.
By cropping the spectra to windows around particular spectral lines, prior knowledge about the wavelengths of important spectral features can be exploited to isolate particular spectral classes with high accuracy in a matter of seconds.
Although dimensionality reduction is a fully unsupervised technique, it can also be used in a supervised way, allowing additional spectra to be classified by visualising their similarity to the rest of the dataset.
With upcoming surveys promising tens of millions of spectra, this work highlights the potential for unsupervised techniques as efficient means of classification and dataset visualisation.
\end{abstract}

@@

White dwarfs (WDs) are the final state of the $\gtrsim97$ per cent of stars with a zero-age mass of less than $9$--$12\,\mathrm{M}_\odot$ \citep{lauffer18, althaus10, althaus21}.
Without internal nuclear fusion, these stellar cinders gradually cool on well-characterised timescales, making them important tracers of the evolution and assembly of the Galaxy (e.g., \citealt{winget87, tremblay14}).
The spectrum of a white dwarf usually deviates from that of a black body due to absorbing species in its atmosphere.
With a typical mass of $\sim0.6\,\mathrm{M}_\odot$ and radius of $\sim \mathrm{R}_E$, the extreme gravitational field gives WDs a strongly stratified structure.
Metals diffuse down through WD atmospheres on timescales much shorter than their cooling age \citep{schatzman45, paquette86, koester09, wyatt14}, leaving thin atmospheres dominated by H and/or He.
Most WD spectra show absorption features due to these light elements, and a WD is classified respectively as DA or DB if spectral features of H or He are present.
As these WDs cool below particular temperatures, atomic transitions cease to be excited ($\lesssim5000\,\text{K}$ for H; $\lesssim 11000\,\text{K}$ for He) and the absorption lines gradually fade away, giving a featureless spectrum (DC).
In certain WDs, the convection zone reaches deep into the interior and dredges up carbon from the inert core, imprinting molecular \ce{C2} Swan bands in the spectrum (DQ; \citealt{fontaine84, koester82, koester20, blouin23}).
In other cases, instabilities in remnant planetary systems can cause planetesimals to be tidally disintegrated and accreted by WDs, lending metal absorption features to the spectrum \citep{bonsor11, frewen14, mustill18, maldonado20}.
These so-called \textit{polluted} WDs (DZ) provide rare insights into the bulk composition and geology of rocky exoplanets.

Following the release of the DESI EDR \citep{desiedr}, the spectra of 3673 WD candidates were visually inspected and classified by \citet{manser24}, spectroscopically confirming 2706 WDs, of which 1400 had not been previously observed spectroscopically.
The higher resolution of DESI ($R\sim2500$--$5000$) enables identification of spectral features not visible in lower-resolution spectra: of the 152 polluted WDs identified, 121 were newly discovered or previously not classified as polluted.
However, visual classification requires significant amounts of expert time, and substantial increases in target numbers in the future ($\sim70\,000$ WD candidates from the DESI first data release; \citealt{cooper23}) all but necessitates the use of automated methods.

One class of methods for automatically classifying and analysing WD spectra that has achieved success is supervised machine learning (for example, \citealt{yang20, tan23, garciazamora23, vincent23, vincent24}).
However, a drawback of supervised machine learning techniques is the reliance on the training set used to train models.
The `imbalanced learning problem' describes the well-documented difficulties of training supervised classifiers on datasets with large class imbalances (e.g., \citealt{he09, johnson19}).
Such class imbalances are inevitable in WD datasets as some spectral types are naturally much rarer than others.
Additionally, incorrect labels in the training set can severely limit the accuracy of the resulting model (e.g., \citealt{frenay14}).
Unsupervised procedures such as dimensionality reduction do not rely on an external training set; they merely reveal structures present within the dataset itself.

In this work, we apply dimensionality reduction -- specifically, $t$SNE -- to intermediate-resolution white dwarf spectra from the DESI EDR, outlining the method's success in characterising WD spectra in an unsupervised way.
We emphasise that these methods could equally be applied to sets of main-sequence stars, quasars, galaxies, or any other set of sources targeted by the aforementioned spectroscopic surveys.
The paper proceeds as follows.
In Section~\ref{sec:methods}, we describe the dataset and dimensionality reduction in more detail, as well as important preprocessing considerations.
In Section~\ref{sec:results}, we demonstrate this method's ability to identify clusters and sequences in the DESI EDR WD catalogue.
We also outline a way to incorporate prior knowledge about the locations of distinctive spectral lines, and a method of using DR in a supervised manner to classify spectra external to the catalogue.
Section~\ref{sec:discussion} discusses the application of this method as an aid in the classification of large spectroscopic WD datasets.
Section~\ref{sec:conclusions} summarises our work.

@@

The DESI EDR contains $N=3673$ WD candidates from the catalogue compiled by \citet{gentilefusillo19} where the exposures include a median signal-to-noise ratio $>0.5$ in at least one of DESI's three spectral arms \citep{manser24}.
These publicly available\footnote{
    \url{https://data.desi.lbl.gov/public/index.html}
} exposures were obtained and restacked.
The spectra span $3600$--$9824\,\text{\AA}$, with a wavelength spacing of $\Delta \lambda = 0.8\,\text{\AA}$, giving $D=7781$ flux bins for each spectrum\footnote{
    Note that the full-width-half-maximum line width of the DESI spectrographs is $\approx1.8\,\text{\AA}$ \citep{abareshi22}; these data are therefore slightly oversampled.
}.

@@

A spectrum can be represented by a high-dimensional vector, with each component of the vector corresponding to the flux in a given wavelength bin.
Spectra produced by the DESI pipeline have 7781 bins, so they can be mapped one-to-one to 7781-dimensional vectors.
These vectors are not uniformly-distributed in $\mathbb{R}^{7781}$;
spectra with shared spectral features are clustered together, as their vector components corresponding to wavelength bins near shared spectral features will take similar values.
For example, the 3703rd component corresponds to a wavelength of $6562.4\,\text{\AA}$, very close to the H$\,\alpha$ line.
The (normalised) spectra of DA WDs will have lower values in their 3703rd component than spectra without any H absorption features.
DAs will hence -- in this component, and by extension other components -- be close together.
Additionally, in the absence of artefacts or large noise, adjacent components of a given spectrum are likely to have similar values.
The components of spectra are thus highly intercorrelated.
This inherent structure in the data means that each point in the dataset lies on or close to a submanifold of the data space with much lower dimensionality.
DR techniques seek to find a low- (usually two-) dimensional representation of the structure of the high-dimensional dataset.

@@

Before applying DR, it is of paramount importance to preprocess the spectra.
The spectra show a range of absolute scales: the median fluxes span three orders of magnitude.
A na\:{i}ve application of DR to the raw data would lead to two identical WDs at different distances (and hence different brightnesses) to be embedded far away from each other.
As we are interested in the intrinsic properties of the WDs in this sample, rather than their distance from Earth, the spectra must be normalised before DR is applied.
Additionally, some spectra show large spikes, due to cosmic rays or detector artefacts (see Fig.~\ref{fig:preprocessing}).
Interpreted as vectors, this corresponds to some components of the vector being very large, and as a result the vector would be very far away in data space from where it would be without the artefacts.

The preprocessing we apply to the data is illustrated in Fig.~\ref{fig:preprocessing}.
Artefacts usually correspond to pixels with very low signal-to-noise: pixels with $\mathrm{S}/\mathrm{N}<0.2$ are linearly interpolated.
Fig.~\ref{fig:preprocessing} shows that major, though not all, artefacts are removed by this step.
Overzealous removal of artefacts could remove genuine spectral features, such as the sharp emission lines of cataclysmic variables (see Section~\ref{sec:CVs}): we found that the interpolation described improved the effectiveness of the subsequent dimensionality reduction as compared to, say, median boxcar smoothing.
Following the removal of artefacts, the spectra are normalised to zero mean and unit variance.

@@

In its simplest form, dimensionality reduction treats all components of the vectors equally.
However, some wavelengths are of more significance than others, namely those of atomic absorption and emission features.
To `focus' the technique on particular spectral lines, we crop the spectrum to an appropriate window before applying DR.
This additional preprocessing step is explored in Section~\ref{sec:zoom}.

@@

Dimensionality reduction was applied to the DESI EDR WD catalogue, using $t$SNE.
The dimensionality reduction takes an average of $5.5\,\text{s}$ on an Intel $3.60\,\text{GHz}$ CPU.
The resulting embedding is shown in Fig.~\ref{fig:full_spectra}(a).
Each point in this embedding corresponds to an individual spectrum, and
several clusters and sequences appear.
There is a long V-shaped sequence stretching from the left, to the bottom, to the upper right of the embedding; it turns off into a cluster towards the centre right.
A shorter secondary sequence extends from the top of the embedding towards the centre.
Several strings and clumps are found in between and around these two sequences.

The nature of these structures is elucidated by Fig.~\ref{fig:full_spectra}(b) and (c), showing respectively the primary spectral class (acc. \citealt{manser24}) and effective temperature (acc. H-atmosphere model fitting of \citealt{gentilefusillo19}).
Several interesting features include:
\begin{itemize}
\item
The V-shaped sequence consists of overwhelmingly DA white dwarfs;
\item
The secondary sequence in the upper part of the frame contains DBs, DZs, DQs, and DCs;
\item
The WDs in the DA sequence have been sorted approximately according to effective temperature, with hotter WDs towards the left of the sequence\footnote{
    This trend is also apparent in the secondary sequence, but because many of these are DBs and DQs, and the temperatures shown here are fitted assuming H-atmosphere models, the trend is less meaningful.
    However, the trend is borne out in the fitted temperatures of He-atmosphere models.
};
\item
DZs are found in various places\footnote{
    \citet{kao24} also find that DZs with different characteristics (primarily $\Teff$) are scattered by dimensionality reduction into multiple regions.
}: (i) an island near to the secondary sequence, (ii) towards the cool end of the DA sequence, (iii) amongst the secondary sequence;
\item
Extragalactic sources (`$\times$' markers) are mostly located in strings and small clumps around the secondary sequence;
\item 
Subdwarfs (`$+$' markers) are largely found towards the hot end of the DA sequence;
\item
The majority of main-sequence stars (black star icons) are found towards the centre right, though their cluster merges somewhat with the cool end of the DA sequence.
A few are also scattered elsewhere.
\end{itemize}

@@

Although the spectral classes are known here, owing to the visual inspection campaign of \citet{manser24}, DR would nonetheless provide a valuable analysis if the classes were \textit{a priori} unknown, as will be the case with upcoming spectroscopic surveys.
For example, it is well-known that approximately 80 per cent of WDs are DAs.
As such, when applying DR to a WD dataset one would implicitly identify the largest cluster as corresponding to mostly DAs.

That DR is capable of separating different classes of WD into groups in this way is ultimately due to similarities between spectra of the same spectral class, and differences between different classes.
Expressed as vectors, DA spectra (say) are much closer to vectors of other DA spectra of similar H abundance and effective temperature, than to DB spectra.
Note that this would not be true if the spectra had not been preprocessed (see Section~\ref{sec:preprocessing}): closer WDs would have brighter spectra than more distant WDs of the same spectral type, and hence spectral vectors with larger magnitudes.
The spectral vectors would therefore be very far away from each other, and hence embedded far away from each other, despite being similar astrophysical objects.
The same would be true of spectra with very large artefacts.
The preprocessing procedure is thus crucial in removing any irrelevant aspects of the data.

The H-atmosphere model fitting of \citet{gentilefusillo19} suggests that the V-shaped sequence transitions smoothly from $\Teff \gtrsim 80000\,\text{K}$ through to $\Teff \lesssim 4000\,\text{K}$.
We see that the variation in the tilt of the spectral black-body continuum has been converted by DR to variation along this H-atmosphere sequence.
The hottest WD in this sequence has a very similar spectrum to the second-hottest, so the two are embedded close together; the second-hottest has a similar spectrum to the third-hottest; etc.
Anomalies of intermediate temperature at opposite ends of the sequence (see Fig.~\ref{fig:full_spectra}(c)) are main-sequence stars or other objects that have been erroneously assigned an ill-fitting WD temperature.
The fact that $\Teff$ ``spans'' the embedding identifies this parameter as the one that describes a spectrum to first order: the parameter which primarily determines the spectrum is the temperature.
If one were not interested in $\Teff$ and did not desire to distinguish between DAs of different temperature, one could include preprocessing steps to erase any temperature information in the spectra (e.g., dividing by the best-fitting black body) before applying DR.
Unable to pick up on this information, DR would then arrange the embedding according to second-order parameters, such as hydrogen abundance.
In such a scenario, the spectrum of a very hot DA would appear similar to that of a very cool DA, and the sequence would thus be collapsed in the resulting embedding.

The transitions undergone by WDs as they gradually cool are also borne out in the spectral classifications of the embedding (Fig.~\ref{fig:full_spectra}(b)), though some subtleties emerge as a consequence of the classification system itself.
As the temperature of a DA falls below around $\Teff\approx 5000\,\text{K}$, hydrogen transitions are no longer excited, the Balmer lines gradually fade away, and the DA becomes a DC.
However, the transition is gradual: the Balmer lines do not suddenly disappear below some sharp temperature cutoff.
As such, the distinction between DA and DC is a fuzzy one, and whether a cool hydrogen-atmosphere WD spectrum is classified as DA or DC depends whether a human classifier can subjectively identify H features against noise.
This task is not trivial, but we argue that it is ultimately arbitrary: there is negligible physical difference between a DA with very weak Balmer lines, and a hydrogen-atmosphere DC.

@@

Similar trends are borne out in the secondary sequence, which shows DBs at the top of the embedding transitioning into a mixture of DCs and DQs.
The transition from DB to DC has a very similar physical origin to the DA-to-DC transition discussed in the previous paragraph, but at higher temperature, as seen in Fig.~\ref{fig:full_spectra}(c).
The DC transition temperatures appear roughly in line with theoretical predictions of $\approx 5000\,\text{K}$ and $\approx 11000\,\text{K}$ for H- and He-dominated atmospheres respectively.
Aside from forming a DC, a cooling He-atmosphere WD may begin dredging carbon from the core, giving rise to a DQ.
As this cooling occurs, He transitions fail to be excited and the He lines of a DB fade away.
As with Balmer lines, He lines and Swan bands can be arbitrarily weak, creating a degree of blurriness between cool DBs, DQs, and DCs.
That these three spectral types are all projected together in the secondary sequence at the top of the embedding reflects the smooth transition between them.
Indeed, Fig.~\ref{fig:full_spectra}(c) shows that the DCs in this part of the embedding have temperatures\footnote{
    Although the temperatures shown in Fig.~\ref{fig:full_spectra}(c) are based on H-atmosphere models, the He-atmosphere model temperatures show the same results.
} well in excess of the DA-to-DC transition temperature of $\approx 5000\,\text{K}$.
These DCs therefore must have He-dominated atmospheres; if they were H-dominated then Balmer features would be visible at these temperatures.
These observations therefore suggest that the two main features of the embedding correspond to H- and He-dominated atmospheres.

Briefly, we discuss the presence of a small number of DAs in the cool end of the He-atmosphere sequence (see Fig.~\ref{fig:full_spectra}(b), just above centre), of which there are 43.
Of these, 24 are classified by \citet{manser24} as DAH, DAe, DAP, or some combination of these secondary classifications.
These may have been found in the He-atmosphere sequence as the Swan bands of the DQs are mimicked by the Zeeman-split Balmer features of a DAH, or emission features of a DAe, or other unidentified features in DAPs.
The remaining 19 DAs here are either pure DAs, DABs, DAZs, or some combination of these.
These DAs are invariably warm ($\Teff \gtrsim 7000\,\text{K}$) and have weak Balmer lines, making them appear similar to DCs of similar temperatures.
It is thus understandable that these DAs are projected near to these warm (thus He-atmosphere) DCs.

It is somewhat surprising that there is a small gap between the two sequences.
Based on the observations outlined above, this region should be populated with DCs of $\approx7000\,\text{K}$, but in fact there are very similar DCs of around this temperature on either side.
Furthermore, it is difficult to explain why there is a clump of DCs at $\approx 6500\,\text{K}$ (which must therefore have He-dominated atmospheres) that are embedded together with the DAs of a similar temperature.
Visual inspection shows that some (though not all) of the DAs in this region have very weak Balmer features, perhaps justifying their proximity to DCs of a similar temperature for reasons discussed above.
It is possible that the gap between this region and the secondary sequence containing the rest of the He-atmosphere WDs is simply due to poor sampling around this temperature, and that with the larger number of objects that will be provided by the full releases of upcoming spectroscopic surveys, this gap will be filled in.
However, the lack of a clear explanation for the sparsity of this region is an undesirable symptom of non-linear DR methods, as opposed to simpler linear methods such as PCA.

The automated nature of DR can protect against human error in the visual classification of WD spectra.
Fig.~\ref{fig:DZisland} shows a zoom-in to the top of the embedding around an island populated mostly by WDs with primary classification DZ.
According to the extensive visual classifications of \citet{manser24}, this island also contains a DBZ (J133305.34+325400.11) and a DC (J160711.86+532157.65).
These objects appear to have been grouped together owing to the presence of Ca~H and K lines in all the spectra, including the putative DC.
The presence of Ca absorption lines in the DC's spectrum suggests that this WD is in fact a DZ; indeed \citet{kleinman13} classify this source as a DZ based on its SDSS DR7 spectrum.
The bottom panel of Figure~\ref{fig:DZisland} shows the absorption lines to be quite weak, and would probably not have been noticeable without zooming in.
While large visual classification campaigns are reliable, this illustrates that errors are nonetheless possible, and that such errors can be easily identified with automated methods such as DR.

@@

As mentioned in Section~\ref{sec:preprocessing}, DR treats all components of a vector -- i.e.\ all wavelengths -- equally, but wavelength ranges that include spectral features are of particular interest.
In this subsection we explore the use of cropping the spectrum, to a window around some particular spectral line, to see if DR can better separate individual spectral classes from the rest.

@@

It is clear that dimensionality reduction alone is not capable of classifying the spectra of all WD candidates.
Though the technique is quickly able to inform the classification of a large fraction of the dataset, particularly when focused on a wavelength range around some distinctive spectral features, we outline drawbacks to DR in this subsection.
